---
output:
  rmarkdown::github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
knitr::opts_chunk$set(fig.width=14, fig.height=8, dpi=300) 
```

# nurah

<!-- badges: start -->
[![Project Status: WIP – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)
[![Codecov test coverage](https://codecov.io/gh/OJWatson/nurah/graph/badge.svg)](https://app.codecov.io/gh/OJWatson/nurah)
[![R-CMD-check](https://github.com/OJWatson/nurah/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/OJWatson/nurah/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`nurah` is an R package designed as a research compendium to evaluate the feasibility of estimating mortality in humanitarian crises. It leverages statistical modelling, directed acyclic graphs (DAGs), and simulations to address data availability challenges, producing robust and actionable mortality estimates.

## Overview

`nurah` facilitates:

1. Defining causal structures for mortality estimation using DAGs.
2. Simulating crisis-affected population data based on defined DAGs.
3. Estimating indirect mortality through Bayesian hierarchical models informed by the causal DAG.

## Installation

Install the latest version from GitHub:

```r
# install.packages("devtools")
devtools::install_github("OJWatson/nurah")
```

## Usage

### Defining a DAG

Start by defining your causal structure with predictors and an outcome (mortality):

```{r}
library(nurah)

nodes <- c("Nutrition", "\"Food insecurity\"", "Mortality")
edges <- data.frame(
  from = c("\"Food insecurity\"", "Nutrition"),
  to = c("Nutrition", "Mortality")
)

dag <- define_dag(nodes, edges)
visualise_dag(dag)
```

### Specifying Parameters

You can add specific parameter values (effect sizes and lags) to your DAG:

```{r}
parameters <- data.frame(
  from = c("\"Food insecurity\"", "Nutrition"),
  to = c("Nutrition", "Mortality"),
  effect_size = c(-0.5, 1.2),
  lag = c(30, 0)
)

dag <- define_dag(nodes, edges, parameters)
```

### Using Checchi et al. (2017) DAG

A core inspiration for `nurah` is the DAG from Checchi et al. (2017). You can quickly use this predefined DAG:

```{r}
dag <- checchi_2017_dag()
visualise_dag(dag, label_size = 4.5)
```

```{r}
# Dummy parameters for demonstration
parms <- dummy_checchi_2017_parameters()
dag <- checchi_2017_dag(parameters = parms)
head(parms)
```

### Simulating Data

Use your DAG to simulate crisis data with defined spatial and temporal structures:

N.B. WARNING - Currently this does not set any values in your DAG if the parent 
nodes have no values! (TODO List)

```r
sim_data <- simulate_crisis_data(
  dag,
  start_date = "2020-01-01",
  n_periods = 6,
  resolution = "month",
  spatial_structure = data.frame(
    country = "X", region = c("A","B"), district = c("A1","B1")
  ),
  initial_population = 10000,
  noise_level = 0.8
)
```

### Estimating Indirect Mortality

Fit a Bayesian hierarchical model to estimate indirect mortality using simulated data:

```r
fit <- estimate_indirect_mortality(
  data = sim_data,
  dag = dag,
  outcome = "Population mortality",
  spatial_levels = c("region", "district"),
  iter = 1000,
  chains = 2
)

summary(fit)
```

## Getting Help

For issues, feature requests, or support, please open an issue on GitHub:

- [GitHub Issues](https://github.com/OJWatson/nurah/issues)

## References

- Checchi et al. (2017). ["Public health information in crisis-affected populations"](https://journals.plos.org/plosmedicine/article?id=10.1371/journal.pmed.1002181). *PLOS Medicine.*
